---
title: "Survival Analysis - Hurricane Readiness Project"
author: "David Kim"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Library & Data

```{r}
library(survival)
library(foreign)
library(ggplot2)
library(survminer)
library(rms)
library(flexsurv)
library(dplyr)
library(tidyverse)
library(ciTools)
library(here)
library(visreg)
library(cmprsk)
library(reticulate)
```

```{r}
library(aws.s3)

# finding the AWS S3 bucket
bucket_exists(
  bucket = "s3://survival2024/",
  region = "us-east-1"
)

# connecting to the bucket
files <- get_bucket_df(
  bucket = "s3://survival2024/",
  region = "us-east-1",
  max = 20000
) %>%
  as_tibble()

# downloading files in the bucket
save_object(
  object = "hurricane.csv",
  bucket = "s3://survival2024/",
  region = "us-east-1",
  file = "hurricane"
)

# hurricane df
df <- read.csv("hurricane")

df
```

```{r}
# View(df)
dim(df)
colnames(df)
```

### 1) Imputation of Missing Values

##### Decision: We're going to treat NA values of h1 to h48 columns with 0.

```{r}
# relocating columns for easier imputation
df <- df |>
  relocate(survive, hour, reason, reason2, .after = elevation)
```

```{r}
# NA values check
any(is.na(df))
```

```{r}
# selecting columns that are h1 to h48
cols_to_impute <- 13:ncol(df)

# looping through the selected columns and replacing NA with 0
for (i in cols_to_impute) {
  df[, i][is.na(df[, i])] <- 0
}

# view the result
View(df)
```

```{r}
# sanity check
any(is.na(df))
```

### 2) Creating a flag for motor failure event

```{r}
# flag motor is when pump fails due to motor failure
df$flag_motor <- ifelse(df$survive == 0 & df$reason == 2, 1, 0)

# moving the column for visibility
df <- df |>
  relocate(flag_motor, .after = survive)

View(df)
```

### 3) Cox PH Model with Time-independent variables

```{r}
# cox proportional hazards model with all time-fixed predictor variables
df.coxph.full <- coxph(Surv(hour, flag_motor) ~ age + backup + bridgecrane + elevation + gear + servo + slope + trashrack, data = df)

summary(df.coxph.full)
```

#### 3.1) Treating trashrack

```{r}
table(df$trashrack, df$flag_motor)
```

In our dataset, all 112 pumps that failed due to motor failure occurred in the absence of trashrack.

This means the variable trashrack in our dataset perfectly predicts the pump failure due to motor failure.

##### Decision: We will leave out trashrack in our model & address it separately

This was done to allow our model to be generalizable to other datasets

We can recommend the committee to 1) look for this type of pattern in other datasets 2) and if pattern is present in our datasets, then suggest thinking about conducting a pilot experiment where upgrade on trashrack is implemented as a solution to addressing motor failure of pumps

```{r}
# without trashrack
df.coxph.full <- coxph(Surv(hour, flag_motor) ~ age + backup + bridgecrane + elevation + gear + servo + slope, data = df)

summary(df.coxph.full)
```

#### 3.2) Checking Linearity Assumption

For continuous variables, we need to check if they satisfy the linearity assumption of Cox PH model

We will use Martingale residual plots to check linearity of predictors

```{r}
# Martingale residual plots for checking linearity assumption in predictors
survminer::ggcoxfunctional(df.coxph.full,data=df)
```

##### Decision: We will bin the continuous variables to satisfy the linearity assumption of predictors

#### 3.3) Binning Continuous Variables

##### Binning Age

We will bin the continuous variables while trying to avoid quasi-complete or complete separation

```{r}
# range of age
sort(unique(df$age))
```

```{r}
# distribution of age
hist(df$age, breaks = 30, main = "Distribution of Age", xlab = "Age")
```

```{r}
# binning age into 2 buckets
# (5-7), (7+)

df <- df |>
  mutate(
    age_binned = cut(age, 
                     breaks = c(-Inf, 7, Inf),
                     labels = c("(5-7)", "(7+)"))
  )
```

```{r}
# checking the distribution of binned age variable
df |> 
    count(age_binned)
```

```{r}
# checking for separation in binned age variable
df |> 
    count(age_binned, flag_motor) |> 
    pivot_wider(names_from = flag_motor, 
                values_from = n)
```

##### Binning Elevation

```{r}
sort(unique(df$elevation))
```

```{r}
hist(df$elevation, breaks = 30, main = "Distribution of Elevation", xlab = "Elevation")
```

```{r}
# binning elevation into 2 buckets
# (2-3), (3+)

df <- df |>
  mutate(
          elevation_binned = cut(elevation,
                           breaks = c(-Inf, 3, Inf),
                           labels = c("(2-3)", "(3+)"))
  )
```

```{r}
# checking the distribution of binned elevation variable
df |> 
    count(elevation_binned)
```

```{r}
# checking for separation in binned elevation variable
df |> 
    count(elevation_binned, flag_motor) |> 
    pivot_wider(names_from = flag_motor, 
                values_from = n)
```

##### Binning Slope

```{r}
sort(unique(df$slope))
```

```{r}
hist(df$slope, breaks = 30, main = "Distribution of Slope", xlab = "Slope")
```

```{r}
# binning slope into 2 buckets
# (0-3), (3+)

df <- df |>
  mutate(
      slope_binned = cut(slope,
                       breaks = c(-Inf, 2, Inf),
                       labels = c("(0-2)", "(2+)"))
  
      )
```

```{r}
# checking the distribution of binned slope variable
df |> 
    count(slope_binned)
```

```{r}
# checking for separation in binned slope variable
df |> 
    count(slope_binned, flag_motor) |> 
    pivot_wider(names_from = flag_motor, 
                values_from = n)
```

#### 3.4) Cox PH Model with Linearity Assumption Satisfied

```{r}
df.coxph.full.lin <- coxph(Surv(hour, flag_motor) ~ age_binned + backup + bridgecrane + elevation_binned + gear + servo + slope_binned, data = df)

summary(df.coxph.full.lin)
```

#### 3.5) Checking for Constant Proportionality of Hazards in the Cox PH Model

This means that the predictor variables should not show interaction with time.

Schoenfeld hypothesis test can be performed on the residuals of the Cox PH model.

The null hypothesis of the Schoenfeld test essentially states that the predictor variable shows no statistically significant interaction with time

The alternative hypothesis of the Schoenfeld test states the predictor variable shows statistically significant interaction with time

Therefore, we want to see high p-values, so that we can fail to reject the null hypothesis

```{r}
df.coxph.full.lin.zph <- cox.zph(df.coxph.full.lin)

df.coxph.full.lin.zph
```

##### Passed!

```{r}
ggcoxzph(df.coxph.full.lin.zph)
```

### 4) Backward Elimination Selection

Committee uses a significance level of 0.03 & backward elimination selection method

##### Decision: We use Backward elimination selection with the significance level of 0.03

```{r}
# significance level
alpha = 0.03

# null model
df.coxph.null <- coxph(Surv(hour, flag_motor) ~ 1, data = df)

# backward elimination with significance level (alpha)
df.coxph.bs <- step(df.coxph.full.lin,
                   scope = list(lower = df.coxph.null,
                                upper = df.coxph.full.lin),
                   direction = "backward", 
                   k = qchisq(alpha, 1,lower.tail = FALSE)
                   ) 
```

```{r}
summary(df.coxph.bs)
```

##### Other Backward Elimination Selection Methods

```{r}
# Backward Elimination with AIC
df.coxph.bs.AIC <- step(df.coxph.full.lin,
                   scope = list(lower = df.coxph.null,
                                upper = df.coxph.full.lin),
                   direction = "backward", 
                   k = 2
                   ) 
```

```{r}
summary(df.coxph.bs.AIC)
```

```{r}
# Backward Elimination with BIC
df.coxph.bs.BIC <- step(df.coxph.full.lin,
                   scope = list(lower = df.coxph.null,
                                upper = df.coxph.full.lin),
                   direction = "backward", 
                   k = log(nrow(df))
                   ) 
```

#### 4.1) P-value table of Significant Variables

```{r}
ranked_pval_tbl <- as.data.frame(summary(df.coxph.bs)$coefficients) |>
  tibble::rownames_to_column("variable") |>
  arrange(`Pr(>|z|)`)

ranked_pval_tbl
```

#### 4.2) Interpretation of Effects of the Most Significant Variable

Our most significant variable is the age variable, but we binned this variable into 2 buckets: (5-7) and (7+)

By binning the age variable, we have to recognize that we sacrificed interpretability for model validity.

```{r}
100*(0.2418283-1)
```

##### Interpretation

"Based on our Cox PH model, age was the strongest predictor of motor failure hazard in pumps. However, it should be noted that we decided to bin the age variable for model validity at the expense of interpretability. Therefore, our interpretation is limited to stating that, while holding all other factors constant, motors that are 7 years or older have approximately 76% lower hazard compared to motors that are 5 to 7 years old."

## Cox Regression Model with Time-Dependent Variable

### 0) Assumptions

We treated NA as 0

h1-h48: "Accuracy of pump status not guaranteed to be error free" –\> We believed what they put down

### 1) Counting Process Structure

```{r}
# 1) selecting only relevant columns
# taking columns OUT (-c)
df_cp_base <- df |>
    select(-c(reason, reason2, bridgecrane, gear, trashrack, elevation, backup, survive))

df_cp_base <- df_cp_base |>
    select(-c(age, servo, slope, elevation_binned))

# creating index
df_cp_base$index = seq_len(nrow(df_cp_base))

df_cp_base <- df_cp_base |>
    relocate(index, .before = flag_motor)

df_cp_base <- df_cp_base |>
    relocate(age_binned, slope_binned, .after = flag_motor)

any(is.na(df_cp_base))
View(df_cp_base)
```

```{r}
# 2) pivot h1:h48 to long format
df_cp <- df_cp_base |>
      pivot_longer(
    cols = matches("^h\\d"),    
    names_to = "stop", 
    values_to = "motor_switch"
  )

View(df_cp)
```

```{r}
# 3) convert 'stop' (e.g., "h1", "h2") to numeric (1, 2)
df_cp <- df_cp |>
    mutate(stop = as.numeric(str_remove(stop, "h")))

View(df_cp)
```

```{r}
# 4) only keep up to the hour of failure/censoring
df_cp <- df_cp |>
    filter(stop <= hour)

View(df_cp)
```

```{r}
# 5) create 'Start' column
df_cp <- df_cp |>
    mutate(start = stop - 1) |>
    relocate(start, .before = stop)

View(df_cp)
```

```{r}
# 6) create Consecutive Running Hour counter
df_cp <- df_cp |>
  group_by(index) |>
  mutate(
    consecutive_running_hours = accumulate(motor_switch, ~ ifelse(.y == 1, .x + .y, 0), .init = 0)[-1]
  ) |>
  ungroup()

View(df_cp)
```

```{r}
# 7) create over_consec_12hrs_flag
df_cp <- df_cp |>
  group_by(index) |>
    mutate(
        over_consec_12hrs_flag = if_else(consecutive_running_hours >= 12, 1, 0)
    ) |>
    ungroup()

View(df_cp)
```

```{r}
# 8) create 'flag_motor_failed' column
df_cp <- df_cp |>
  group_by(index) |>
  mutate(
    flag_motor_failed = ifelse(flag_motor == 0, 0,   # censored gets 0 all the way
                        ifelse(stop == hour, 1, 0))  # event gets 1 when stop
  ) |>
  ungroup() |>
    relocate(flag_motor_failed, .after = stop)

View(df_cp)
```

### 2) Adding Time-Dependent variable to the Cox PH Model

```{r}
df.coxph.time_dep <- coxph(Surv(start, stop, flag_motor_failed) ~ age_binned + slope_binned + over_consec_12hrs_flag,
                           data = df_cp)

summary(df.coxph.time_dep)
```

#### Interpretation

We created a time-dependent variable that indicates if the motor was running for over 12 hours at the time of motor pump failure.
We added this variable to the Cox PH model with significant predictor variables that met the requirements of linearity and time-independence (shown earlier).
Inside this model, this time dependent variable was shown to be not statistically significant in its impact, meaning that the over-drive of motor for 12+ consecutive hours didn't seem to play a significant role in increasing the hazard of motor pump failures.
